Multiresolution Neural Field Estimation

### 2011-02-28 Multiresolution_framework/ltx/manuscript.tex
added stoping criteria to EM
fixed the label of Fig3

\section{State and Parameter Estimation}
The state-space representation allows the use of the expectation maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two-part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters (M-step) and the posterior distribution of the field weights (E-step). The M-step finds an estimate of $\boldsymbol\theta$ that maximizes the lower bound, $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$, of the complete-data log-likelihood,  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$. The E-step uses the estimate of $\boldsymbol\theta$, $\boldsymbol\theta'$ and the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} to estimate the posterior state sequence. In this paper we use a standard EM methodology. Therefore, we only describe aspects of the algorithm that are specific to MRA neural field equations due to the space limitations. The bound on the log-likelihood is a quadratic in the form of
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
with a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables are calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$, and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}
\dean{Number of iterations and stopping criteria?}

@@

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 %\includegraphics[width=0.5\textwidth]{./Graph/RMSE.pdf}
\includegraphics[scale=1]{./Graph/RMSE.pdf}
 \caption{RMSE of the estimated field at different spatial resolution.}
 \label{fig:RMSE}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{\dean{the label a) is missing from the figure.} Spatial field at one time instant for different spatial resolutions, estimated and true fields are shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{Histograms of the parameter estimates over 100 realizations. In each of the subplots
the solid lines show the actual parameter values and the dotted lines show the means of the estimated parameters. Left panel: The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. Right panel: The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$.}
 \label{fig:ParametersHistogram}
 \end{figure}

>>>>

\section{State and Parameter Estimation}
The state-space representation allows the use of the expectation maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two-part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters (M-step) and the posterior distribution of the field weights (E-step). The M-step finds an estimate of $\boldsymbol\theta$ that maximizes the lower bound, $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$, of the complete-data log-likelihood,  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$. The E-step uses the estimate of $\boldsymbol\theta$, $\boldsymbol\theta'$ and the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} to estimate the posterior state sequence. In this paper we use a standard EM methodology. Therefore, we only describe aspects of the algorithm that are specific to MRA neural field equations due to the space limitations. The bound on the log-likelihood is a quadratic in the form of
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
with a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in Eqs. \ref{eq:upsilon0}, \ref{eq:upsilon1}, and \ref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables are calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$, and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).
\end{align}
\dean{Number of iterations and stopping criteria?}The stopping criterion is usually associated  with  either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}.

@@

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
\begin{figure}[!h]
 \centering
 \includegraphics[scale=1]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations generated by the kernel in (a). The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 %\includegraphics[width=0.5\textwidth]{./Graph/RMSE.pdf}
\includegraphics[scale=1]{./Graph/RMSE.pdf}
 \caption{RMSE of the estimated field at different spatial resolution.}
 \label{fig:RMSE}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Field.pdf}
 \caption{Spatial field at one time instant for different spatial resolutions, estimated and true fields are shown by dashed and solid lines respectively; (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{Histograms of the parameter estimates over 100 realizations. In each of the subplots
the solid lines show the actual parameter values and the dotted lines show the means of the estimated parameters. Left panel: The histogram of the central excitatory connectivity kernel basis function parameter estimates, $\theta_0$. Right panel: The histogram of the surround inhibition connectivity kernel basis function parameter estimates, $\theta_1$.}
 \label{fig:ParametersHistogram}
 \end{figure}
